---
title: "Lab 10: Distances and the Border Zone"
subtitle: 'Ecosystem Science and Sustainability 330'
author:
  - name: Andie Hall
    email: andiehal@colostate.edu
format: html
execute:
  echo: true
---

## Library Code

```{r}
# Spatial Data Science
library(tidyverse)
library(sf)
library(units)
library(flextable)

# Data Access
library(AOI)

# Visualization
library(gghighlight)
library(ggrepel)
library(knitr)

# Install required packages (only run once)
# remotes::install_github("mikejohnson51/AOI")
# remotes::install_github("ropensci/USAboundaries")
# remotes::install_github("ropensci/USAboundariesData")
# remotes::install_github("ropenscilabs/rnaturalearthdata")
```

### Question 1: Ingesting / building sf objects from R packages and CSVs.

```{r}
# 1.1 - Define a Projection
eqdc <- '+proj=eqdc +lat_0=40 +lon_0=-96 +lat_1=20 +lat_2=60 +x_0=0 +y_0=0 +datum=NAD83 +units=m +no_defs'

# 1.2 - Get USA state boundaries

state_boundaries <- aoi_get(state = 'conus') |> 
  st_transform(eqdc)

us_states <- rnaturalearth::ne_states(country = "United States of America", returnclass = "sf") |>
  filter(!name %in% c("Alaska", "Hawaii", "Puerto Rico"))

us_states <- st_transform(us_states, crs = eqdc)

# 1.3 - Get country boundaries for Mexico, the United States of America, and Canada
countries <- aoi_get(country = c("MX", "CA", "USA"))

# 1.4 - Get city locations from the CSV file and removing AK, HI, and PR
uscities <- read_csv("data/simplemaps_uscities_basicv1.90/uscities.csv")

spatial_uscities <- st_as_sf(uscities, coords = c("lng", "lat"), crs = 4326) |>
  st_transform(eqdc) |>
  st_filter(us_states)

# Glimpse structure and plot
glimpse(spatial_uscities)
plot(spatial_uscities$geometry, pch = 16, cex = 0.1, main = "US Cities")
```

### Question 2: Manipulating geometries and coordinate systems

#### 2.1 - Distance to USA Border (coastline or national) (km)

```{r}
usa_union <- st_union(us_states)
usa_polygon <- st_sf(geometry = usa_union)
usa_border <- usa_polygon |> 
  st_cast("MULTILINESTRING")

plot(usa_border$geometry, col = "blue", lwd = 2, main = "USA Border")

spatial_uscities <- spatial_uscities |> 
  mutate(distance_to_border_km = as.numeric(st_distance(geometry, usa_border)) / 1000)

top5_border_cities <- spatial_uscities |> 
  arrange(desc(distance_to_border_km)) |> 
  select(city, state_name, distance_to_border_km) |> 
  slice_head(n = 5)

flextable(top5_border_cities |> 
            rename(
              `City` = city,
              `State` = state_name,
              `Distance to Border (km)` = distance_to_border_km
            ))
```

#### 2.2 - Distance to States (km)

```{r}

state_borders <- st_boundary(us_states)

state_borders_combined <- st_union(state_borders)
state_borders_combined <- st_cast(state_borders_combined, "MULTILINESTRING")

conus_state_ids <- us_states$abbrev

cities_proj <- st_transform(spatial_uscities, 5070)
borders_proj <- st_transform(state_borders_combined, 5070)

distances <- st_distance(cities_proj, borders_proj)
spatial_uscities$dist_to_us_border_m <- as.numeric(distances)

top5_border <- spatial_uscities |>
  select(city, state_id, dist_to_us_border_m) |>
  mutate(distance_km = round(dist_to_us_border_m / 1000, 2)) |>
  arrange(desc(distance_km)) |>
  slice(1:5)

flextable::flextable(top5_border)
```

#### 2.3 - Distance to Mexico (km) 

```{r}

mexico <- countries |> filter(admin == "Mexico")
mexico_border <- st_boundary(mexico)
mexico_border <- st_cast(mexico_border, "MULTILINESTRING")
mexico_proj <- st_transform(mexico_border, 5070)

mexico_distances <- st_distance(cities_proj, mexico_proj)
spatial_uscities$dist_to_mexico_m <- as.numeric(mexico_distances)

mexico_topcities <- spatial_uscities |>
  select(city, state_id, dist_to_mexico_m) |>
  mutate(distance_km = round(dist_to_mexico_m / 1000, 2)) |>
  arrange(desc(distance_km)) |>
  slice(1:5)

flextable::flextable(mexico_topcities)

```


#### 2.4 - Distance to Canada (km) 

```{r}
canada <- countries |> filter(admin == "Canada")
canada_border <- st_boundary(canada)
canada_border <- st_cast(canada_border, "MULTILINESTRING")
canada_proj <- st_transform(canada_border, 5070)

canada_distances <- st_distance(cities_proj, canada_proj)
spatial_uscities$dist_to_canada_m <- as.numeric(canada_distances)

canada_top5 <- spatial_uscities |>
  select(city, state_id, dist_to_canada_m) |>
  mutate(distance_km = round(dist_to_canada_m / 1000, 2)) |>
  arrange(desc(distance_km)) |>
  slice(1:5)

flextable::flextable(canada_top5)
```


### Question 3: Calculating distances

### 3.1 - Data

```{r}
conus_outline <- st_union(us_states)

top10_cities <- read_csv("data/simplemaps_uscities_basicv1.90/uscities.csv", show_col_types = FALSE) |>
  arrange(desc(population))|>
  slice(1:10)

top10_cities_sf <- st_as_sf(top10_cities, coords = c("lng", "lat"), crs = 4326)

ggplot() +
  geom_sf(data = countries, fill = "gray") +
  geom_sf(data = conus_outline, lty = "dashed") +
  geom_sf(data = us_states) +
  geom_sf(data = top10_cities_sf) +
  ggrepel::geom_label_repel(
    data = top10_cities_sf,
    aes(label = city, geometry = geometry),
    stat = "sf_coordinates") +
  theme_minimal()

```

#### 3.2 - City Distance from the Border
```{r}

ggplot() +
  geom_sf(data = countries, fill = "gray") +
  geom_sf(data = conus_outline, lty = "dashed") +
  geom_sf(data = us_states) +
  geom_sf(data = top5_border_cities) +
  ggrepel::geom_label_repel(
    data = top5_border_cities,
    aes(label = city, geometry = geometry),
    stat = "sf_coordinates") +
  theme_minimal()

```

#### 3.3 - City Distance from Nearest State

```{r}

ggplot() +
  geom_sf(data = countries, fill = "gray") +
  geom_sf(data = conus_outline, lty = "dashed") +
  geom_sf(data = us_states) +
  geom_sf(data = top5_border) +
  ggrepel::geom_label_repel(
    data = top5_border,
    aes(label = city, geometry = geometry),
    stat = "sf_coordinates") +
  theme_minimal()

```

#### 3.4 - Equidistance boundary from Mexico and Canada

```{r}

spatial_uscities$dist_to_mexico <- as.numeric(st_distance(cities_proj, mexico_proj))
spatial_uscities$dist_to_canada <- as.numeric(st_distance(cities_proj, canada_proj))

spatial_uscities$border_diff <- abs(spatial_uscities$dist_to_mexico - spatial_uscities$dist_to_canada)

equal_distance <- spatial_uscities |>
  filter(border_diff <= 100000)

top5_equal <- equal_distance |>
  arrange(desc(population)) |>
  slice(1:5)

top5_equal_sf <- st_as_sf(top5_equal, coords = c("lng", "lat"), crs = 4326)

ggplot() +
  geom_sf(data = countries, fill = "gray") +
  geom_sf(data = conus_outline, lty = "dashed") +
  geom_sf(data = us_states) +
  geom_sf(data = top5_equal_sf) +
  ggrepel::geom_label_repel(
    data = top5_equal_sf,
    aes(label = city, geometry = geometry),
    stat = "sf_coordinates") +
  theme_void()

```

### Question 4: Building maps using ggplot

### 4.1 - Quantifing Border Zone

```{r}
border_area <- 160934

border_zone <- spatial_uscities |>
  filter(dist_to_mexico <= border_area | dist_to_canada <= border_area)

total_cities <- nrow(uscities)
total_population <- sum(uscities$population, na.rm = TRUE)

border_cities <- nrow(border_zone)
border_population <- sum(border_zone$population, na.rm = TRUE)

border_pct <- ((border_population / total_population) * 100)

summary_tbl <- tibble(
  `Metric` = c("Total cities", "Cities in 100-mile border zone",
               "Total population", "Population in border zone", 
               "Percent of total population in zone"),
  `Value` = c(
    formatC(total_cities, format = "d", big.mark = ","),
    formatC(border_cities, format = "d", big.mark = ","),
    formatC(total_population, format = "d", big.mark = ","),
    formatC(border_population, format = "d", big.mark = ","),
    sprintf("%.1f%%", border_pct)))

flextable(summary_tbl)

```

#### 4.2 - Mapping Border Zone

```{r}
danger_zone_cities <- spatial_uscities |>
  filter(dist_to_mexico_m <= 160934 | dist_to_canada_m <= 160934) |>
  mutate(dist_to_border_m = pmin(dist_to_mexico_m, dist_to_canada_m))

top10_danger <- danger_zone_cities |>
  arrange(desc(population)) |>
  slice(1:10)

top10_danger_sf <- st_as_sf(top10_danger, coords = c("lng", "lat"), crs = 4326)

ggplot() +
  geom_sf(data = countries, fill = NA, color = "black", lwd = 0.5) +
  geom_sf(data = danger_zone_cities, aes(color = dist_to_border_m), size = 1.2) +
  ggrepel::geom_label_repel(
    data = top10_danger,
    aes(label = city, geometry = geometry),
    stat = "sf_coordinates",
    min.segment.length = 0
  ) +
  scale_color_gradient(low = "darkred", high = "orange", name = "Distance to Border (m)") +
  theme_void() +
  theme(legend.position = "bottom")

```

#### 4.3 - Most Populous Danger City for Each State

```{r}
label_cities <- danger_zone_cities |>
  group_by(state_id) |>
  filter(population == max(population, na.rm = TRUE)) |>
  slice(1) |> 
  ungroup()

ggplot() +
  geom_sf(data = countries, fill = NA, color = "black", lwd = 0.5) +
  geom_sf(data = danger_zone_cities, aes(color = dist_to_border_m), size = 1.2) +
  ggrepel::geom_label_repel(
    data = label_cities,
    aes(label = city, geometry = geometry),
    stat = "sf_coordinates",
    min.segment.length = 0
  ) +
  scale_color_gradient(low = "darkred", high = "orange", name = "Distance to Border (m)") +
  theme_void() +
  theme(legend.position = "bottom")
```

